---
title: 'BIOSTAT 203B: Homework #4, Exercise #4'
author: "Chad Pickering"
date: "3/16/2018"
output: html_document
---

##### Connect, etc.:

```{r, warning=FALSE, message=FALSE}
if (!"gridExtra" %in% rownames(installed.packages())){
  install.packages("gridExtra", repos="http://cran.rstudio.com/")
}

if (!"viridis" %in% rownames(installed.packages())){
  install.packages("viridis", repos="http://cran.rstudio.com/")
}

library(sparklyr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(gridExtra)
library(viridis)

Sys.setenv(SPARK_HOME="/usr/lib/spark")
config <- spark_config()
sc <- spark_connect(master = "yarn-client", config = config)
flights_tbl <- tbl(sc, 'flights')
airports_tbl <- tbl(sc, 'airports')
airlines_tbl <- tbl(sc, 'airlines')
```

### 4. Build a predictive model for the arrival delay (`arrdelay`) of flights flying from LAX. You are allowed to use a maximum of 5 predictors.  

#### Model selection:  

I am subsetting the dataset to include only flights from 2003 to 2007 - predictions will be more indicative of current conditions. 

```{r, echo=FALSE}
model_data <- flights_tbl %>%
  filter(origin == "LAX") %>%  
  filter(!is.na(arrdelay) & !is.na(depdelay) & 
           !is.na(actualelapsedtime) & !is.na(distance)) %>%
  filter(year >= 2003 & year <= 2007) %>%
  filter(depdelay > 15 & depdelay < 240) %>%
  filter(arrdelay > -60 & arrdelay < 360) %>%
  left_join(airlines_tbl, by = c("uniquecarrier" = "code")) %>%
  select(arrdelay, uniquecarrier, description,
         depdelay, actualelapsedtime, distance)
```

##### Models attempted with >= 2003 and no other filters: 

depdelay R2=0.8334, RMSE=11.81
depdelay + actualelap R2=0.834, RMSE=11.79
depdelay + distance R2=0.8338, RMSE=11.79
depdelay + dayofweek R2=0.8334, RMSE=11.81
depdelay + arrtime R2=0.8334, RMSE=11.81
depdelay + uniquecarrier R2=0.8352, RMSE=11.74
depdelay + actualelap + distance R2=0.9294, RMSE=7.686
depdelay + uniquecarrier + actualelap + distance R2=0.9349, RMSE=7.381

To avoid overfitting with more predictors, I am choosing the last mentioned with range 2003-2007.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Partition the data into training and validation sets
model_partition <- model_data %>% 
  sdf_partition(train = 0.8, valid = 0.2, seed = 5555)

# Fit a linear model
ml1 <- model_partition$train %>%
  ml_linear_regression(arrdelay ~ depdelay + uniquecarrier + 
                         actualelapsedtime + distance)

summary(ml1)
```

#### Model performance:

Here, we generate deciles to assess model performance.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}
model_deciles <- lapply(model_partition, function(x) {
    sdf_predict(ml1, x) %>%
      mutate(decile = ntile(desc(prediction), 10)) %>%
      group_by(decile) %>%
      summarize(arrdelay_mean = mean(arrdelay)) %>%
      select(decile, arrdelay_mean) %>%
      collect()
  })

model_deciles
```

The training vs. validation deciles indicate that the model is quite good.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}
deciles <- rbind(
  data.frame(data = 'train', model_deciles$train),
  data.frame(data = 'valid', model_deciles$valid),
  make.row.names = FALSE
)

deciles %>%
  ggplot(aes(factor(decile), arrdelay_mean, fill = data)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(title = 'Average arrival delay by predicted decile', x = 'Decile', y = 'Minutes')
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
data_2008 <- flights_tbl %>%
  filter(origin == "LAX") %>%  
  filter(!is.na(arrdelay) & !is.na(depdelay) & 
           !is.na(actualelapsedtime) & !is.na(distance)) %>%
  filter(year == 2008) %>%
  filter(depdelay > 15 & depdelay < 240) %>%
  filter(arrdelay > -60 & arrdelay < 360) %>%
  left_join(airlines_tbl, by = c("uniquecarrier" = "code")) %>%
  select(arrdelay, uniquecarrier, description,
         depdelay, actualelapsedtime, distance)
```

The model is biased, but still quite accurate. If predicted mean arrival delay is adjusted downwards by about two to three minutes, this would be a very well-performing model. It is still great, though.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}
carrier <- sdf_predict(ml1, data_2008) %>%
  group_by(description) %>%
  summarize(arrdelay_mean = mean(arrdelay), prediction = mean(prediction), freq = n()) %>%
  filter(freq > 300) %>%
  collect()

carrier
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(carrier, aes(arrdelay_mean, prediction)) + 
  geom_point(alpha = 0.75, color = 'red', shape = 3) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.15, color = 'blue') +
  geom_text(aes(label = substr(description, 1, 20)), size = 2, alpha = 0.75, vjust = -1) +
  labs(title='Average Arrival Delay Forecast', x = 'Actual', y = 'Predicted')
```









